---
title: "MathLab - Lecture 4"
output:   learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
mtcars$am <- factor(mtcars$am, labels = c("automatic", "manual"))
mtcars$vs <- factor(mtcars$vs, labels = c("V", "S"))
mtcars$cyl <- factor(mtcars$cyl)
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
library(ggplot2)
```

## Exploratory Data Analysis

In this lab we will implement our first exploratory analysis in R. We will use the `mtcars` dataset we have been working with throughout this module. As a first starting point it is always good to check the structure of the `dataframe` to ensure the data is stored correctly.

```{r mtcars, exercise = TRUE, exercise.lines = 2}
# Check the structure of the mtcars dataset

```

So we see that the variables `cyl`, `vs`, `am`, `gear` and `carb` are (correctly) stored as factors in `mtcars`. So we can start our exploratory data analysis. A good starting point is the function `summary` which, as the name suggests, produces a summary of the dataset.

```{r summary, exercise = TRUE, exercise.lines =2}
# Use the function summary over the mtcars dataset

```

So we can see that summary reports some summary statistics, as for instance mean and median, for continuous variables, whilst it reports the number of observations in each level for categorical variables. 

Sometimes we may also want to produce these summary statistics by levels of a categorical variable. This can be achieved with the command `by`. Let's consider an example:

 - `by(mtcars, mtcars$am, summary)`
 
The above code produces a `summary` of the dataset `mtcars` by grouping according to the levels of the variable `am`. Now you try!

```{r by, exercise = TRUE, exercise.lines =2}
# Use the function by to report summaries grouped by the variable vs

```

The above functions are included in base R, but of course packages can provide more advanced functions for the purpose of summarizing data. Here is a list:

 - `describe` from the `Hmisc` package
 - `stat.desc` from the `pastecs` package
 - `describe` from the `psych` package
 - `skim` from the `skimr` package
 - `descr` from the `summarytools` package

More details can be found at this [link](https://dabblingwithdata.wordpress.com/2018/01/02/my-favourite-r-package-for-summarising-data/).

## Tables and Barplots

Frequency tables show you how often a given value occurs. To look at a frequency table of the data in R, use the function `table`. The top row of the output is the value, and the bottom row is the frequency of the value. Let's use `table` on the `am` variable of the mtcars data set. 

```{r table1, exercise = TRUE, exercise.lines =2}
# Compute the frequency table of am
```

We can also compute joint tables using two categorical variables. For instance, `table(mtcars$am,mtcars$vs)` constructs a table with the number of observations in every combination of levels. 

```{r table2, exercise = TRUE, exercise.lines = 2}
# Compute the joint table for cyl and am

```

We can also combine the function `table` with `prop.table` to report proportions instead of total counts. For instance, `prop.table(table(mtcars$am))` reports the proportions of cars by transmission.

The command `table` is also the starting point to construct barplots using the function `barplot`. By using `barplot(table(mtcars$am))` we can construct a barplot of the variable `am`.

```{r barplot, exercise = TRUE, exercise.lines =2}
# Construct a barplot of cyl
```

Stacked barplots can be constructed by giving in input tables of two variables as we practiced in the previous exercises.

```{r barplot2, exercise = TRUE, exercise.lines =2}
# Construct a stacked barplot of cyl and am

```

We can of course also customize the plots to make them a bit more appealing visually. Run the code below for you to try.

```{r barplot3, exercise = TRUE, exercise.lines = 3}
counts <- table(mtcars$cyl,mtcars$am)
barplot(counts, beside = TRUE, col = c("blue","green","red"), legend = rownames(counts), ylab = "number of cars", xlab = "Transmission")
```

## Histograms

To visualize continuous data we often want to use the histogram. The function `hist` produces an histogram of a variable. 

```{r hist1, exercise = TRUE, exercise.lines = 2}
# Try constructing an histogram of the variable mpg

```

Two interesting inputs of the function `hist` are:
 
 - `breaks`: this can be given as a number that ``fixes" the number of bars in the histogram
 - `freq`: if `TRUE` it reports counts, if `FALSE` it reports frequencies/proportions.
 
```{r hist2, exercise = TRUE, exercise.lines = 2}
# Modify the previous histogram to have 10 bars and show proportions

```

Plots (not only histograms) can be overall customized. Here are some options:

- `col`: to choose the color (it can be a string or a number);
- `main`: a string to give a title to the plot;
- `xlim`: a vector of length 2 giving the limits of the x-axis;
- `ylim`: as `xlim` but for the y-axis;
- `xlab`: a string to give a label to the x-axis;
- `ylab`: as `xlab` but for the y-axis.

```{r hist3, exercise = TRUE, exercise.lines =2}
# Construct a red histogram of mpg with title Histogram of MPG

```

The function `density` computes the values of the smooth version of the histogram. We can add the density plot to an histogram using the function `lines` (this more generally adds a line to a plot). Have a look at the code below for an example

```{r hist4, exercise = TRUE, exercise.lines = 2}
hist(mtcars$mpg, freq = F, xlim=c(5,40))
lines(density(mtcars$mpg), lwd = 2 , lty =2 , col = "blue")
```

Here we see two additional option of `lines`: `lwd` - the width of the line; `lty` - the type of the line.

## Boxplots

Boxplots can describe the distribution of a continuous random variable. They can be created with the function `boxplot`.

```{r boxplot1, exercise = TRUE, exercise.lines =2}
# Construct a boxplot of mpg
```

As you can see the plot is quite basic, but anyway we will rarely use a boxplot to just describe a continuous variable. In its most powerful use, the boxplot compares a continuous distribution by levels of a categorical variable. For instance, `boxplot(mpg~am, data = mtcars)` produces boxplots of `mpg` for different values of `am`.

```{r boxplot2, exercise = TRUE, exercise.lines = 2}
# Construct a boxplot of mpg by cyl
```

## Correlation

The numerical summary that we introduced to summarize the relationship between two continuous variables is the so-called correlation. The function `cor` computes the correlation between any two (or more) numerical variables. For instance, `cor(mtcars$mpg,mtcars$hp)` computes the correlation between `mpg` and `hp`. 

Of course, we may want to more quickly produce all possible correlations. Let's try and give as input the whole dataframe `mtcars`.

```{r corr, exercise = TRUE, exercise.lines = 2}
# Run the following code
cor(mtcars)
```

It tells us `x must be numeric`. We are giving as input also factors, for which R does not now how to compute the correlation. One solution is to manually select the columns we want to retain. Let's create the correlation between `mpg`, `disp`, `hp`, `drat`, `wt` and `qsec`. These are the 1st, 3rd, 4th, 5th, 6th and 7th column of `mtcars`. I also give you code to more quickly doing this, without having to select the columns manually.

```{r corr1, exercise = TRUE, exercise.lines =6}
# Compute the correlation between the variables mentioned above


# An automatic way to do this
cor(mtcars[sapply(mtcars, is.numeric)])

```

The above code creates a symmetric matrix with all possible pairs of correlations. There are various libraries that can create some more pleasant output, as for instance the `corrplot` package and its `corrplot` function. Here's an example (notice that `corrplot` takes as input a correlation matrix). 

```{r corrplot, exercise = TRUE, exercise.lines = 2}
library(corrplot)
corrplot(cor(mtcars[sapply(mtcars, is.numeric)]))
```

There are many options that can be set for this plot. Have a look at the help for more info. Other representations of correlations can be found at this [link](http://www.sthda.com/english/wiki/correlation-matrix-a-quick-start-guide-to-analyze-format-and-visualize-a-correlation-matrix-using-r-software).

## Scatterplots

The scatterplot is the graphical representation of the relationship between two variables. This can be produced with the function `plot`. 

```{r scat, exercise = TRUE, exercise.lines = 6}
# Scatterplot between mpg and hp
plot(mtcars$hp, mtcars$mpg)

# Create a scatterplot between disp and mpg
```

Again the output of `plot` is very basic and we can change it using some of the options we have already seen (e.g. `xlab`, `ylab`, `main`, `col`). But there are others:

- `pch`: a number to choose the plotting symbol;
- `cex`: to change the size of the plotting symbol;

```{r scat1, exercise = TRUE, exercise.lines = 3}
# Create a scatterplot between mpg and hp, with pch = 20 with blue symbols
```

We can also color code observations depending on a third categorical variable. The option `col` takes as input either a string or a number, which is associated to colors. The following example shows you how to update the plot above to color points depending on `am`.

```{r scat2, exercise = TRUE, exercise.lines =2}
plot(mtcars$hp,mtcars$mpg, pch=20, col = as.numeric(mtcars$am))
legend("topright", legend = c("automatic","manual"),pch=20, col= c(1,2))
```

Lines can be added to a plot using the `abline` command. You can look at the help page to find additional details.

If we want to produce the scatterplots between multiple variables at the same time, we can use the function `pairs`. The following code produces the scatterplots between the first four variables of `mtcars` (this is so that the output is actually visible)

```{r pairs, exercise = TRUE, exercise.lines =2}
pairs(mtcars[,1:4])
```

For the categorical variable `cyl`, it produces some sort of boxplot. If we want we can exclude the non-numeric variables as done for computing the correlation. 

Furthermore, the function `ggpairs` from the `GGally` package creates a fancier pairs plot. Let's try it!

```{r pairs2, exercise = TRUE, exercise.lines = 2}
library(GGally)
ggpairs(mtcars[,1:4])
```

This plot is a lot more informative and actually reports a lot of the plots we have seen in the previous sections. The function has a large number of options that can be set, you can find them typing `?ggpairs` once the `GGally` package is installed and loaded.

## ggplot

The previous commands allow to very quickly produce graphs, but their costumization to make them look pretty is often painful. There is widely-used data visualization package in R called `ggplot2` which creates beautiful data visualizations: actually many of the plots you showed in the forum were created with `ggplot2`. It is impossible to cover the many features of this package shortly, and this tutorial is really to give you an intuition of how things work. By searching in google, you will find a huge amount of tutorials of various complexity/comprehensiveness. 

In order to use `ggplot` you first need to install it once, and then load it at the beggining of any R session (just like any other package). In this tutorial it is loaded for you already.

ggplot constructs any plot starting from a `dataframe`. In a same plot con only appear things that are stored or are computed from variables appearing in the same `dataframe`. Very very often this is actually all we need, but sometimes we need some extra flexibility, and this is why it is useful to know a little bit of the base R graphics.

Any plot created with `ggplot2` starts with the function `ggplot` and the dataframe we are interested in. Let's have a look

```{r ggplot, exercise = TRUE, exercise.lines =2}
ggplot(mtcars)
```

The above code creates an empty canvas. Anything drawn over this canvas has to derive from the `mtcars` dataset. Next we specify aesthetic mappings, which specify how you want to map variables to visual aspects. In this case we are simply mapping the `disp` and `mpg` variables to the x- and y-axes.

```{r ggplot1, exercise = TRUE, exercise.lines =2}
ggplot(mtcars, aes(x = disp, y = mpg))
```
The above code creates an empty canvas where now we specified that on the x-axis there is `disp` and on the y-axis there is `mpg`.

Then we can add new layers that are geometric objects which will show up on the plot. In this case we add `geom_point` to add a layer with points (dot) elements as the geometric shapes to represent the data.

```{r ggplot2, exercise = TRUE, exercise.lines  = 2}
ggplot(mtcars, aes(x = disp, y = mpg)) + geom_point()
```

### Aesthetic Mappings

The aesthetic mappings take properties of the data and use them to influence visual characteristics, such as position, color, size, shape, or transparency. Each visual characteristic can thus encode an aspect of the data and be used to convey information.

All aesthetics for a plot are specified in the `aes()` function call (below you will see that each geom layer can have its own aes specification). For example, we can add a mapping from the number of cylinders of the cars to a color characteristic:

```{r ggplot3, exercise = TRUE, exercise.lines = 2}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

### Specifying Geometric Shapes
Building on these basics, `ggplot2` can be used to build almost any kind of plot you may want. The most obvious distinction between plots is what geometric objects (geoms) they include. `ggplot2` supports a number of different types of geoms, including:

 - `geom_point()` for scatterplots;
 - `geom_line()` for drawing lines;
 - `geom_smooth()` for drawing smoothed lines;
 - `geom_bar()` for barplots;
 - `geom_boxplot()` for boxplots;
 - `geom_histogram()` for histograms;

Let's have a look at the following examples.

```{r ggplot4, exercise = TRUE, exercise.lines =9}
# A boxplot
ggplot(mtcars, aes(x=cyl,y=mpg)) + geom_boxplot(col="black", fill="blue")

# A histogram 
ggplot(mtcars, aes(x=mpg)) + geom_histogram(color="darkblue", fill="lightblue", bins = 5)

# A barplot
ggplot(mtcars, aes(x=cyl)) + geom_bar(fill="steelblue")
```

What makes this really powerful is that you can add multiple geometries to a plot, thus allowing you to create complex graphics showing multiple aspects of your data.

```{r ggplot5, exercise = TRUE, exercise.lines = 3}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() +
  geom_smooth()
```

We can specify certain aes within each geom, which allows us to only show certain characteristics for that specificy layer. Have a look at the next two plots and see if you can understand their difference.

```{r ggplot6, exercise = TRUE, exercise.lines = 7}
ggplot(mtcars, aes(x = disp, y = mpg, color = am)) +
  geom_point() +
  geom_smooth(se = FALSE, method = "lm")

ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point(aes(color = am)) +
  geom_smooth(se = FALSE, method = "lm")
```
