---
title: "MathLab - Lab 2"
output:   learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
mtcars$am <- factor(mtcars$am, labels = c("automatic", "manual"))
mtcars$vs <- factor(mtcars$vs, labels = c("V", "S"))
library(ggplot2)
```

## Exploratory Data Analysis

In this lab we will implement our first exploratory analysis in R. We will use the `mtcars` dataset we have been already working with.

The aim of an exploratory analysis is to have an overview of the data of interest and to discern interesting patterns. Such patterns may then be used to guide the modeling we will learn in the next lectures.

Steps of an exploratory analysis may include:

 - checking that the data is stored correctly in the software (variable types, missing values etc.)
 
 - create numerical summaries of the data
 
 - produce graphical representations of the data
 
As a first starting point it is always good to check the structure of the `dataframe` to ensure the data is stored correctly using the function `str`.

```{r mtcars, exercise = TRUE, exercise.lines = 2}
# Check the structure of the mtcars dataset

```

So we see that the variables  `vs` and `am` are (correctly) stored as factors in `mtcars`. So we can start our exploratory data analysis. 


## Numerical Summaries

Often one of the first steps in an analysis is to produce numerical summaries of the data. This can be achieved with the function `summary` which, as the name suggests, produces a summary of the dataset.

```{r summary, exercise = TRUE, exercise.lines =2}
# Use the function summary over the mtcars dataset

```

So we can see that summary reports some summary statistics, as for instance mean and median, for continuous variables, whilst it reports the number of observations in each level for categorical variables. 

Sometimes we may also want to produce these summary statistics by levels of a categorical variable. This can be achieved with the command `by`. Let's consider an example:

 - `by(mtcars, mtcars$am, summary)`
 
The above code produces a `summary` of the dataset `mtcars` by grouping according to the levels of the variable `am`. Now you try!

```{r by, exercise = TRUE, exercise.lines =2}
# Use the function by to report summaries grouped by the variable vs

```

The above functions are included in base R, but of course packages can provide more advanced functions for the purpose of summarizing data. Here is a list:

 - `describe` from the `Hmisc` package
 - `stat.desc` from the `pastecs` package
 - `describe` from the `psych` package
 - `skim` from the `skimr` package
 - `descr` from the `summarytools` package

More details can be found at this [link](https://dabblingwithdata.wordpress.com/2018/01/02/my-favourite-r-package-for-summarising-data/).


## Frequency Tables

Frequency tables show you how often a given value occurs. To look at a frequency table of the data in R, use the function `table`. The top row of the output is the value, and the bottom row is the frequency of the value. Let's use `table` on the `am` variable of the mtcars data set. 

```{r table1, exercise = TRUE, exercise.lines =2}
# Compute the frequency table of am
```

We can also compute joint tables using two categorical variables. For instance, `table(mtcars$am,mtcars$vs)` constructs a table with the number of observations in every combination of levels. 

```{r table2, exercise = TRUE, exercise.lines = 2}
# Compute the joint table for cyl and am

```

We can also combine the function `table` with `prop.table` to report proportions instead of total counts. For instance, `prop.table(table(mtcars$am))` reports the proportions of cars by transmission.


## Correlation

The numerical summary that we introduced to summarize the relationship between two quantitative variables is the so-called correlation. The function `cor` computes the correlation between any two (or more) quantitative variables. For instance, `cor(mtcars$mpg,mtcars$hp)` computes the correlation between `mpg` and `hp`. 

Of course, we may want to more quickly produce all possible correlations.  All variables in `mtcars` are quantitative expect for `vs` and `am`. We can use the following code to compute the correlation between all other variables.

```{r corr, exercise = TRUE, exercise.lines = 2}
# Run the following code
cor(mtcars[,-c(8,9)])
```


The above code creates a symmetric matrix with all possible pairs of correlations. There are various libraries that can create some more pleasant output, as for instance the `corrplot` package and its `corrplot` function. Here's an example (notice that `corrplot` takes as input a correlation matrix). 

```{r corrplot, exercise = TRUE, exercise.lines = 2}
library(corrplot)
corrplot(cor(mtcars[,-c(8,9)]))
```

There are many options that can be set for this plot. Have a look at the help for more info. Other representations of correlations can be found at this [link](http://www.sthda.com/english/wiki/correlation-matrix-a-quick-start-guide-to-analyze-format-and-visualize-a-correlation-matrix-using-r-software).


## Standard R Graphics

R provides a wide array of functions to produce data visualizations. They are easy to use, they quickly produce an output, but they usually require a lot of effort so that they are customized to be pleasantly looking. Personally, I use them when I want to get a quick visualization, but I then use so more advanced package if I have to share these plots. 

We will not spend a lot of time discussing these and I just give you a list of functions that can be used:

 - **barplots**: the function `barplot` can be used which takes as input a `table`, e.g. `barplot(table(mtcars$cyl))`.
 
 - **histograms**: the function `hist` takes as input a vector of data and outputs an histogram, e.g. `hist(mtcars$mpg)`.
 
 - **density plots**: the function `density` construct the density of the data. This can be plotted on its own, e.g. `plot(density(mtcars$mpg))` or added to an histogram, e.g. `lines(density(mtcars$mpg))`.

 - **boxplots**: the function `boxplot` creates a boxplot. If the interest is in describing a categorical vs a continuous variable one can use e.g. `boxplot(mtcars$mpg ~ mtcars$cyl)`.
 
 - **scatterplots**: the function `plot` with inputs two vectors of data creates a scatterplot, e.g. `plot(mtcars$mpg,mtcars$hp)`. The function `pairs` which takes as input a dataframe creates a scatterplot between every pair of variables.
 
 
### Examples 

For you to see the capabilities of basic R, try and run the codes below and look at the output.

```{r barplot, exercise = TRUE, exercise.lines =2}
# Simple Barplot
barplot(table(mtcars$cyl))
```


```{r barplot3, exercise = TRUE, exercise.lines = 3}
# Bivariate Barplot with Customization
counts <- table(mtcars$cyl,mtcars$am)
barplot(counts, beside = TRUE, col = c("blue","green","red"), legend = rownames(counts), ylab = "number of cars", xlab = "Transmission")
```

```{r hist1, exercise = TRUE, exercise.lines = 2}
# A simple histogram
hist(mtcars$mpg)
```


```{r hist4, exercise = TRUE, exercise.lines = 2}
# A fancier histogram with density line
hist(mtcars$mpg, freq = F, xlim=c(5,40))
lines(density(mtcars$mpg), lwd = 2 , lty =2 , col = "blue")
```

```{r boxplot2, exercise = TRUE, exercise.lines = 2}
# A boxplot of mpg by cyl
boxplot(mtcars$mpg~mtcars$cyl)
```

```{r scat, exercise = TRUE, exercise.lines = 2}
# Scatterplot between mpg and hp
plot(mtcars$hp, mtcars$mpg)
```


```{r scat2, exercise = TRUE, exercise.lines =3}
# A Scatterplot with color and customization
plot(mtcars$hp,mtcars$mpg, pch=20, col = as.numeric(mtcars$am))
legend("topright", legend = c("automatic","manual"),pch=20, col= c(1,2))
```


```{r pairs, exercise = TRUE, exercise.lines =2}
# A pairs plot
pairs(mtcars[,1:4])
```

## An Overview Plot

The function `ggpairs` from the `GGally` package creates a fancier pairs plot. Let's try it!

```{r pairs2, exercise = TRUE, exercise.lines = 2}
library(GGally)
ggpairs(mtcars[,1:8])
```

This plot is a lot more informative and actually reports a lot of the plots we have seen in the previous sections. The function has a large number of options that can be set, you can find them typing `?ggpairs` once the `GGally` package is installed and loaded.

## ggplot

The previous commands allow to very quickly produce graphs, but their costumization to make them look pretty is often painful. There is widely-used data visualization package in R called `ggplot2` which creates beautiful data visualizations: actually many of the plots you showed in the forum were created with `ggplot2`. It is impossible to cover the many features of this package shortly, and this tutorial is really to give you an intuition of how things work. By searching in google, you will find a huge amount of tutorials of various complexity/comprehensiveness. 

In order to use `ggplot` you first need to install it once, and then load it at the beggining of any R session (just like any other package). In this tutorial it is loaded for you already.

ggplot constructs any plot starting from a `dataframe`. In a same plot con only appear things that are stored or are computed from variables appearing in the same `dataframe`. Very very often this is actually all we need, but sometimes we need some extra flexibility, and this is why it is useful to know a little bit of the base R graphics.

Any plot created with `ggplot2` starts with the function `ggplot` and the dataframe we are interested in. Let's have a look

```{r ggplot, exercise = TRUE, exercise.lines =2}
ggplot(mtcars)
```

The above code creates an empty canvas. Anything drawn over this canvas has to derive from the `mtcars` dataset. Next we specify aesthetic mappings, which specify how you want to map variables to visual aspects. In this case we are simply mapping the `disp` and `mpg` variables to the x- and y-axes.

```{r ggplot1, exercise = TRUE, exercise.lines =2}
ggplot(mtcars, aes(x = disp, y = mpg))
```
The above code creates an empty canvas where now we specified that on the x-axis there is `disp` and on the y-axis there is `mpg`.

Then we can add new layers that are geometric objects which will show up on the plot. In this case we add `geom_point` to add a layer with points (dot) elements as the geometric shapes to represent the data.

```{r ggplot2, exercise = TRUE, exercise.lines  = 2}
ggplot(mtcars, aes(x = disp, y = mpg)) + geom_point()
```

### Aesthetic Mappings

The aesthetic mappings take properties of the data and use them to influence visual characteristics, such as position, color, size, shape, or transparency. Each visual characteristic can thus encode an aspect of the data and be used to convey information.

All aesthetics for a plot are specified in the `aes()` function call (below you will see that each geom layer can have its own aes specification). For example, we can add a mapping from the number of cylinders of the cars to a color characteristic:

```{r ggplot3, exercise = TRUE, exercise.lines = 2}
ggplot(mtcars, aes(x = disp, y = mpg, color = am)) +
  geom_point()
```

### Specifying Geometric Shapes
Building on these basics, `ggplot2` can be used to build almost any kind of plot you may want. The most obvious distinction between plots is what geometric objects (geoms) they include. `ggplot2` supports a number of different types of geoms, including:

 - `geom_point()` for scatterplots;
 - `geom_line()` for drawing lines;
 - `geom_smooth()` for drawing smoothed lines;
 - `geom_bar()` for barplots;
 - `geom_boxplot()` for boxplots;
 - `geom_histogram()` for histograms;

Let's have a look at the following examples.

```{r ggplot4, exercise = TRUE, exercise.lines =9}
# A boxplot
ggplot(mtcars, aes(x=cyl,y=mpg)) + geom_boxplot(col="black", fill="blue")

# A histogram 
ggplot(mtcars, aes(x=mpg)) + geom_histogram(color="darkblue", fill="lightblue", bins = 5)

# A barplot
ggplot(mtcars, aes(x=cyl)) + geom_bar(fill="steelblue")
```

What makes this really powerful is that you can add multiple geometries to a plot, thus allowing you to create complex graphics showing multiple aspects of your data.

```{r ggplot5, exercise = TRUE, exercise.lines = 3}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() +
  geom_smooth()
```

We can specify certain aes within each geom, which allows us to only show certain characteristics for that specificy layer. Have a look at the next two plots and see if you can understand their difference.

```{r ggplot6, exercise = TRUE, exercise.lines = 7}
ggplot(mtcars, aes(x = disp, y = mpg, color = am)) +
  geom_point() +
  geom_smooth(se = FALSE, method = "lm")

ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point(aes(color = am)) +
  geom_smooth(se = FALSE, method = "lm")
```
