---
title: "MathLab - Lecture 2"
output:   learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Vectors
Vectors are one-dimension arrays that can hold one same type of data. In other words, a vector is a simple tool to store data. For example, you can store your daily gains and losses in the casinos.

In R, you create a vector with the combine function c(). You place the vector elements separated by a comma between the brackets. For example:

 - `numeric_vector <- c(1, 2, 3)`
 - `character_vector <- c("a", "b", "c")`
 - `boolean_vector <- c(TRUE, FALSE)`

Complete the code such that boolean_vector contains the three elements: `TRUE`, `FALSE` and `TRUE` (in that order).

```{r vec, exercise=TRUE, exercise.lines = 5}
numeric_vector <- c(1, 10, 49)
character_vector <- c("a", "b", "c")

# Complete the code for 'boolean_vector'

```

<div id="vec-hint">
Assign `c(TRUE,FALSE,TRUE)` to the variable `boolean_vector` with the `<-` operator. You can also use the shorter notations, `T` and `F` for `TRUE` and `FALSE` respectively.
</div>

## Manipulating dataframes

We saw in last class that dataframes are a matrix of data, where observations are stored on the rows and the relevant variables are stored on the columns. Different columns can store data of a different class.

We can select rows and columns of a dataframe using the square brackets `[]`. For instance, if we wanted the entry in the second row and third row of the dataframe `mtcars` we can use the code `mtcars[2,3]`. If we want a full row or full column we have to omit the first or second number within the square brackets, e.g.

 - `mtcars[2,]` returns the second row of `mtcars`
 - `mtcars[,3]` returns the third row of `mtcars`

We can also select multiple rows/columns of a dataframe by using vectors. For instance `mtcars[c(1,2),3]` returns the first two rows and the third column of `mtcars`.

```{r index, exercise = TRUE, exercise.lines = 6}
# Select the 1st and 10th row of mtcars

# Select the 4th column of mtcars

# Select the 2nd and 4th rows and 3rd and 4th columns of mtcars

```

```{r index-solution}
mtcars[c(1,10),]
mtcars[,4]
mtcars[c(2,4),c(3,4)]
```

### Selecting columns

You can also use the `$` operator to select an entire column from a dataframe. For instance, `mtcars$am` would select the entire `am` column from the dataframe `mtcars`.

```{r dollar, exercise = TRUE, exercise.lines =2}
# Select the column hp from mtcars

```

```{r dollar-solution}
mtcars$hp
```

### Selection by comparison
Sometimes you want to select elements from a dataframe in a more advanced fashion. This is where the use of logical operators may come in handy.

The (logical) comparison operators known to R are: 
 - `<` for less than 
 - `>` for greater than
 - `<=` for less than or equal to 
 - `>=` for greater than or equal to 
 - `==` for equal to each other 
 - `!=` not equal to each other

These can be used within the function `subset` to select rows and column of a dataframe. For example `subset(mtcars, mtcars$hp>100)` creates a dataframe where only observations such that horsepower is bigger than 100 are retained.

```{r subset, exercise = TRUE, exercise.lines = 4}
# Select observations in mtcars with 6 cylinders

# Select observations in mtcars that do not have 4 cylinders

```

```{r subset-hint}
# you may want to use the function subset together with the logical operators == (to check equality) and != (to check difference)
```

```{r subset-solution}
subset(mtcars, mtcars$cyl == 6)
subset(mtcars, mtcars$cyl != 4)
```

We can also impose some more complicated rule by using the `&` logical "and" and the `|` logical or. For instance to return a subset of `mtcars` including cars with a horsepower larger than 100 and four cylinders we can use `subset(mtcars, mtcars$hp >100 & mtcars$cyl == 4)`.

```{r ciao, exercise = TRUE, exercise.lines = 2}
# Select observations in mtcars with 6 cylinders and mpg less than 20

```

```{r ciao-hint}
# Try using subset(mtcars, mtcars$mpg < 20 & mtcars$cyl == 6)
```

### Adding columns 

We can add a column to a dataframe by simply assigning a vector of values to a column having a new name. For instance, suppose we want to compute the sum of `hp` and `mpg` in `mtcars` and we want to store that information in a column called `hp_mpg`. Then we can do this by

`mtcars$hp_mpg <- mtcars$hp + mtcars$hp_mpg`

## Inspecting dataframes
There are several functions you can use to inspect your dataframe. To name a few

 - `head`: this by default prints the first 6 rows of the dataframe
 - `tail`: this by default prints the last 6 rows to the console
 - `str`: this prints the structure of your dataframe
 - `dim`: this by default prints the dimensions, that is, the number of rows and columns of your dataframe
 - `colnames`: this prints the names of the columns of your dataframe
 
```{r insp, exercise = TRUE, exercise.lines = 4}
# Return the first six rows of mtcars

# Print the names of the variables in mtcars

```

```{r insp-solution}
head(mtcars)
colnames(mtcars)
```

### Data structure

Using the `str()` function we can look at the structure of a dataset. The output shows the variable names, their type, and the values of the first observations. This function is particularly important to ensure when loading a dataset into R, the process was correctly done.

```{r str, exercise = TRUE, exercise.lines = 2}
# Return the structure of mtcars

```

```{r str-solution}
str(mtcars)
```


## Functions 

In programming, functions are used to incorporate sets of instructions that we want to use repeatedly. A function is actually a piece of code written to carry out a specified task; it may accept arguments or parameters (or not) and it may return one or more values (or not!).

Let's look at a pre-programmed function in R: `mean`. To consult the R documentation on this function, you can use the following commands:

 - `help(mean)`
 - `?mean`
 
Try these commands out in the console. If you do so, you'll be redirected to www.RDocumentation.org. If you would type this function into you R studio console, a help tab would automatically open in R studio.

There is another way of getting help on a function. For instance, if you want to know which parameters need to be provided, you can use the R function args on the specified function. An example of using args on a function is the following: `args(mean)`

```{r help, exercise=TRUE, exercise.lines = 5}
# ask for help on the mean function


# ask for the arguments used by the mean function

```

<div id="help-hint">
You can ask help about any function by typing `?function`, for instance, `?sd` to get help on the standard deviation. You can do the same for the `mean` function
</div>

### Matching

In the last exercise we made a start with functions. Also, we looked at how we could get help on using functions.

When getting help on the mean function, you saw that it takes an argument x, which is just an arbitrary name for the object that you want to find the mean of. Usually this object will be an R vector. We also saw the .... This is called an elipsis and is used to provide a number of optional arguments to the function.

R can match arguments both by position and by name. Let's say we want to find the mean of a vector called `temperature`. An example of matching by name is the following:

 - `mean(x = temperature)`
 
An example of matching by position is the following:
 
 - `mean(temperature)`

In this exercise, I have provided you with a vector of 5 numbers. There are the grades that you got during the semester.

```{r matching, exercise = TRUE, exercise.lines = 9}
# a grades vector
grades <- c(8.5, 7, 9, 5.5, 6)

# calculate the mean of grades using matching by name


# calculate the mean of grades using matching by position
```

```{r matching-solution}
mean(x = grades)
mean(grades)
```

When we looked at the documentation of `mean`, the documentation showed us the following method:

 - `mean(x, trim = 0, na.rm = FALSE, ...)`

As you can see, both `trim` and `na.rm` have default values. However, `x` doesn't. This makes `x` a required argument. That means that the function mean will throw an error if `x` hasn't been specified. `trim` and `na.rm` are however optional arguments with default values and can be changed or specified by the user.

`na.rm` can be changed by the user if a given vector contains missing values. For instance, if a the aforementioned vector called `temperature` would have missing values, calling mean on it would throw an output of `NA`. If you want the mean function to exclude the `NA` values when calculating the mean, you can specify `na.rm = TRUE`. Let's bring this into practice.

```{r missing, exercise = TRUE, exercise.lines = 9}
# a grades vector
grades <- c(8.5, 7, 9, NA, 6)

# Calculate the mean without removing the missing value


# Calculate the mean removing the missing value
```

```{r missing-hint}
#The key here is specifying the na.rm argument. This can be set to TRUE and FALSE
```

```{r missing-solution}
mean(grades)
mean(grades, na.rm = TRUE)
```

## Importing data

In the last class we learnt that there are two functions to import data `read.table` and `read.csv`. Furthermore we also learnt how to import data using the features provided by R Studio. Let's look at the help for `read.table`.

```{r read, exercise = TRUE, exercise.lines = 2}
# Open the help page of read.table
```

```{r read-solution}
?read.table
```

We see that the function has many inputs, let's highlight the most important ones:

 - `file`: the name of the file you are trying to load, it must include the path to the file if it is not in the working directory;
 
 - `header`: whether or not the file includes names of the variables in the first line;
 
 - `sep`: the character used to separate entries in the file
 
 - `na.strings`: the charachter used to denote a missing observation.

### Investigating the mtcars dataset

Let's look back at the `mtcars` dataset. Let's look at its help page and its structure.

```{r mtcarshelp, exercise = TRUE, exercise.lines = 6}
# Load the help page of mtcars


# Output the structure of mtcars

```

```{r mtcarshelp-solution}
?mtcars
str(mtcars)
```

The variables `vs` and `am` are stored as numerics, but the numbers 0 and 1 used are just labels to denote whether the transmission is automatic or manual, for instance. So for instance if we were to run the code

`mtcars$am[1] + mtcars$am[2]`

which is the sum of the value of `am` for the first and second observations, R would output the number 2. In this setting this does not mean anything.

So perhaps something did not work successfully when this dataset was imported into R, since `vs` and `am` should not be numeric.

## Factors 

Such categorical variables should be stored in R as a `factor`. The function `factor` can be used to recode the variables `am` and `vs` appropriately. This takes three important inputs (you can look at `?factor`):

 - `x` the vector you want to transform into a factor
 - `levels` an optional vector of the unique values (as character strings) that `x` might have taken
 - `labels` an optional character vector of labels for the levels
 
So for instance we can transform the variable `vs` into a factor by `mtcars$vs <- factor(mtcars$vs,labels = c("V","S")`.

```{r am, exercise = TRUE, exercise.lines = 6}
# Transform the variable am of mtcars into a factor


# Report the structure of mtcars

```

```{r am-solution}
mtcars$am <- factor(mtcars$am, labels = c("automatic","manual"))
str(mtcars)
```

So we can see that now `am` is a factor which takes 2 levels called `automatic` and `manual` and that for instance the first three observations are `manual`.

If a variable is stored as a `factor` we can retrieve its levels using the function `levels`.
