---
title: "MathLab - Lecture 2"
output:   learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Vectors
Vectors are one-dimension arrays that can hold one same type of data. In other words, a vector is a simple tool to store data. For example, you can store your daily gains and losses in the casinos.

In R, you create a vector with the combine function `c()`. You place the vector elements separated by a comma between the brackets. For example:

 - `numeric_vector <- c(1, 2, 3)`
 - `character_vector <- c("a", "b", "c")`
 - `boolean_vector <- c(TRUE, FALSE)`

Complete the code such that `boolean_vector` contains the three elements: `TRUE`, `FALSE` and `TRUE` (in that order).

```{r vec, exercise=TRUE, exercise.lines = 5}
numeric_vector <- c(1, 10, 49)
character_vector <- c("a", "b", "c")

# Complete the code for 'boolean_vector'

```

<div id="vec-hint">
Assign `c(TRUE,FALSE,TRUE)` to the variable `boolean_vector` with the `<-` operator. You can also use the shorter notations, `T` and `F` for `TRUE` and `FALSE` respectively.
</div>

